---
title: "Business Intelligence Project"
author: "Trevor Okinda"
date: "27th November 2023"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **BI Project Group Name/ID (if applicable)** | A Rainfall prediction model with Random Trees |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset

### Source: 

The dataset that was used can be downloaded here: *\<provide a link\>*

### Reference:

*\<Cite the dataset here using APA\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

```{r Load datasetr}
library(readr)
WeatherData <- read.csv("weather.csv", colClasses = c(
  MinTemp = "numeric",
  MaxTemp = "numeric",
  Rainfall = "numeric",
  Evaporation = "numeric",
  Sunshine = "numeric",
  WindGustDir = "factor",
  WindGustSpeed = "numeric",
  WindDir9am = "factor",
  WindDir3pm = "factor",
  WindSpeed9am = "numeric",
  WindSpeed3pm = "numeric",
  Humidity9am = "numeric",
  Humidity3pm = "numeric",
  Pressure9am = "numeric",
  Pressure3pm = "numeric",
  Cloud9am = "numeric",
  Cloud3pm = "numeric",
  Temp9am = "numeric",
  Temp3pm = "numeric",
  RainToday = "factor",
  RISK_MM = "numeric",
  RainTomorrow = "factor"
))


# Define levels for categorical columns
wind_dir_levels <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
rain_levels <- c("No", "Yes")

# Update factor columns with levels
WeatherData$WindGustDir <- factor(WeatherData$WindGustDir, levels = wind_dir_levels)
WeatherData$WindDir9am <- factor(WeatherData$WindDir9am, levels = wind_dir_levels)
WeatherData$WindDir3pm <- factor(WeatherData$WindDir3pm, levels = wind_dir_levels)
WeatherData$RainToday <- factor(WeatherData$RainToday, levels = rain_levels)
WeatherData$RainTomorrow <- factor(WeatherData$RainTomorrow, levels = rain_levels)

View(WeatherData)

```
## Categorical columns analysis
```{r}
# List of categorical columns in the dataset
categorical_columns <- c("WindGustDir", "WindDir9am", "WindDir3pm", "RainToday", "RainTomorrow")

# Loop through each categorical column and generate frequency and percentage tables
for (col in categorical_columns) {
  column_freq <- table(WeatherData[[col]])
  
  cat(paste("# Number of instances in (", col, "):\n"))
  class_counts <- cbind(frequency = column_freq, percentage = prop.table(column_freq) * 100)
  print(class_counts)
  cat("\n")
}
```
## Measures of Central Tendency

```{r}
# List of numeric columns in the dataset
numeric_columns <- c(
  "MinTemp", "MaxTemp", "Rainfall", "Evaporation", "Sunshine",
  "WindGustSpeed", "WindSpeed9am", "WindSpeed3pm",
  "Humidity9am", "Humidity3pm",
  "Pressure9am", "Pressure3pm",
  "Cloud9am", "Cloud3pm",
  "Temp9am", "Temp3pm",
  "RISK_MM"
)

# Mode function definition
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Loop through each numeric column and calculate measures of central tendency
for (col in numeric_columns) {
  cat(paste("# Measures of central tendency for", col, ":\n"))
  
  # Mean
  mean_val <- mean(WeatherData[[col]])
  cat("Mean:", mean_val, "\n")
  
  # Median
  median_val <- median(WeatherData[[col]])
  cat("Median:", median_val, "\n")
  
  # Mode (using the Mode function defined below)
  mode_val <- Mode(WeatherData[[col]])
  cat("Mode:", mode_val, "\n\n")
}

summary(WeatherData)
```
## Measures of Distribution
```{r}
# List of numeric columns in the dataset
numeric_columns <- c(
  "MinTemp", "MaxTemp", "Rainfall", "Evaporation", "Sunshine",
  "WindGustSpeed", "WindSpeed9am", "WindSpeed3pm",
  "Humidity9am", "Humidity3pm",
  "Pressure9am", "Pressure3pm",
  "Cloud9am", "Cloud3pm",
  "Temp9am", "Temp3pm",
  "RISK_MM"
)

# Loop through each numeric column and calculate measures of distribution
for (col in numeric_columns) {
  cat(paste("# Measures of Distribution for", col, ":\n"))
  
  # Range
  range_val <- range(WeatherData[[col]])
  cat("Range:", range_val, "\n")
  
  # Variance
  var_val <- var(WeatherData[[col]])
  cat("Variance:", var_val, "\n")
  
  # Standard Deviation
  sd_val <- sd(WeatherData[[col]])
  cat("Standard Deviation:", sd_val, "\n")
  
}
```
## Measures of Relationship
```{r}
# List of numeric columns in the dataset
numeric_columns <- c(
  "MinTemp", "MaxTemp", "Rainfall", "Evaporation", "Sunshine",
  "WindGustSpeed", "WindSpeed9am", "WindSpeed3pm",
  "Humidity9am", "Humidity3pm",
  "Pressure9am", "Pressure3pm",
  "Cloud9am", "Cloud3pm",
  "Temp9am", "Temp3pm",
  "RISK_MM"
)

# Loop through pairs of numeric columns and calculate measures of relationship
for (i in 1:(length(numeric_columns) - 1)) {
  for (j in (i + 1):length(numeric_columns)) {
    col1 <- numeric_columns[i]
    col2 <- numeric_columns[j]
    
    cat(paste("# Measures of Relationship between", col1, "and", col2, ":\n"))
    
    # Correlation
    correlation_val <- cor(WeatherData[[col1]], WeatherData[[col2]])
    cat("Correlation:", correlation_val, "\n")
    
    # Covariance
    covariance_val <- cov(WeatherData[[col1]], WeatherData[[col2]])
    cat("Covariance:", covariance_val, "\n\n")
  }
}
```
## ANOVA
```{r}
# ANOVA for a categorical variable (WindGustDir) and a numeric variable (MaxTemp)
anova_result1 <- aov(MaxTemp ~ WindGustDir, data = WeatherData)

# Display the ANOVA result for the first analysis
cat("# ANOVA Result for MaxTemp and WindGustDir:\n")
print(anova_result1)

# ANOVA for another categorical variable (RainTomorrow) and a numeric variable (MaxTemp)
anova_result2 <- aov(MaxTemp ~ RainTomorrow, data = WeatherData)

# Display the ANOVA result for the second analysis
cat("\n# ANOVA Result for MaxTemp and RainTomorrow:\n")
print(anova_result2)

```
## Basic Visualizations



