---
title: "Business Intelligence Project"
author: "Trevor Okinda"
date: "27th November 2023"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **BI Project Group Name/ID (if applicable)** | A Rainfall prediction model with Random Trees |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset

### Source: 

The dataset that was used can be downloaded here: *\<provide a link\>*

### Reference:

*\<Cite the dataset here using APA\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

```{r Dataset Loader}
library(readr)
WeatherData <- read.csv("weather.csv", colClasses = c(
  MinTemp = "numeric",
  MaxTemp = "numeric",
  Rainfall = "numeric",
  Evaporation = "numeric",
  Sunshine = "numeric",
  WindGustDir = "factor",
  WindGustSpeed = "numeric",
  WindDir9am = "factor",
  WindDir3pm = "factor",
  WindSpeed9am = "numeric",
  WindSpeed3pm = "numeric",
  Humidity9am = "numeric",
  Humidity3pm = "numeric",
  Pressure9am = "numeric",
  Pressure3pm = "numeric",
  Cloud9am = "numeric",
  Cloud3pm = "numeric",
  Temp9am = "numeric",
  Temp3pm = "numeric",
  RainToday = "factor",
  RISK_MM = "numeric",
  RainTomorrow = "factor"
))

# Define levels for categorical columns
wind_dir_levels <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
rain_levels <- c("No", "Yes")

# Update factor columns with levels
WeatherData$WindGustDir <- factor(WeatherData$WindGustDir, levels = wind_dir_levels)
WeatherData$WindDir9am <- factor(WeatherData$WindDir9am, levels = wind_dir_levels)
WeatherData$WindDir3pm <- factor(WeatherData$WindDir3pm, levels = wind_dir_levels)
WeatherData$RainToday <- factor(WeatherData$RainToday, levels = rain_levels)
WeatherData$RainTomorrow <- factor(WeatherData$RainTomorrow, levels = rain_levels)

View(WeatherData)

```

...to be continued